### 2013-06-14 doc/pgas13/abstract.tex
Incorporate Simon's feedback.

% Structured concurrency (scheduler, cooperative multithreading + communication) enables hierarchical flat-combining to be implemented efficiently for a software PGAS shared-memory solution.
The implementation of scalable synchronized data structures is notoriously difficult. Recent work in shared-memory multicores introduced a new synchronization paradigm called \emph{flat combining} that allows many concurrent accessors to efficiently cooperate to reduce contention on shared locks.
In this work we introduce this paradigm to a domain where reducing communication is paramount: distributed memory systems.
We implement a flat combining framework for Grappa, a latency-tolerant PGAS runtime, and show how it can be used to implement synchronized global data structures.
Even using simple locking schemes, we find that these flat-combining data structures scale out to 64 nodes with 2x-100x improvement in throughput. We also demonstrate that this translates to application performance with two simple graph analysis kernels.
The higher communication cost and structured concurrency of Grappa lead to a new form of distributed flat combining that drastically reduces the amount of communication necessary for maintaining global sequential consistency.

>>>>

% Structured concurrency (scheduler, cooperative multithreading + communication) enables hierarchical flat-combining to be implemented efficiently for a software PGAS shared-memory solution.
The implementation of scalable synchronized data structures is notoriously difficult. Recent work in shared-memory multicores introduced a new synchronization paradigm called \emph{flat combining} that allows many concurrent accessors to cooperate efficiently to reduce contention on shared locks.
In this work we introduce this paradigm to a domain where reducing communication is paramount: distributed memory systems.
We implement a flat combining framework for Grappa, a latency-tolerant PGAS runtime, and show how it can be used to implement synchronized global data structures.
Even using simple locking schemes, we find that these flat-combining data structures scale out to 64 nodes with 2x-100x improvement in throughput. We also demonstrate that this translates to application performance via two simple graph analysis kernels.
The higher communication cost and structured concurrency of Grappa lead to a new form of distributed flat combining that drastically reduces the amount of communication necessary for maintaining global sequential consistency.

### 2013-06-14 doc/pgas13/flat-combining.tex
Incorporate Simon's feedback.

The above example of delegation is compelling in itself. However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack, and the requesting threads' records are marked as complete.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.

@@

All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object. To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete. Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}, the task must send a message to the master to acquire the lock. If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All other workers performing operations block at the first message until the lock is released. Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait. However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.

@@

To preserve sequentially consistent semantics for flat-combined operations, they must obey a consistent global order.
To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Stack).
% haven't discussed queue, set, or map yet...
%(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.

@@

In the same way as inserts, lookups can piggy-back on other lookups of the same key.
It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted.
However, this allows the local order in which keys were inserted to be observed.
If this is allowed, then that same order must be respected globally, forcing each batch to ensure atomicity from all other cores' batches. This is prohibitively expensive, so a cheaper solution is chosen: lookups must get their values from the global data structure, but can be performed in parallel in the same way as inserts.

>>>>

The above example of delegation is compelling in itself. However, the crux of prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack, and the requesting threads' records are marked as complete.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.

@@

All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object. To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete. Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}. The task must send a message to the master to acquire the lock. If successful, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All others block at the first message until the lock is released. Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait. However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.

@@

To preserve sequentially consistent semantics for flat-combined operations, they must obey a consistent global order.
To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the \emph{master} core for the Stack).
% haven't discussed queue, set, or map yet...
%(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.

@@

In the same way as inserts, lookups can piggy-back on other lookups of the same key.
It is tempting to allow a lookup to be satisfied locally by inspecting the keys being inserted.
However, this allows the local order in which keys were inserted to be observed.
If this is allowed, then that same order must be respected globally, forcing each batch to commit atomically with respect to other cores' batches. Enforcing this is prohibitively expensive, so a cheaper solution is chosen: if lookups must get their values from the global data structure, then batches can be performed in parallel.

### 2013-06-14 doc/pgas13/grappa.tex
Incorporate Simon's feedback.

\subsection{Global Memory}
In the PGAS style, Grappa has a notion of a global address space which is partitioned across the physical memories of the nodes in a cluster. Each core owns a portion of memory, which is divided among execution stacks for the core's Workers, a core-local heap, and a slice of the global heap.

>>>>

\subsection{Global Memory}
In the PGAS style, Grappa provides a global address space partitioned across the physical memories of the nodes in a cluster. Each core owns a portion of memory, which is divided among execution stacks for the core's Workers, a core-local heap, and a slice of the global heap.

### 2013-06-14 doc/pgas13/intro.tex
Incorporate Simon's feedback.

Globally shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently. It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some serializable order that all threads agree on (known as \emph{sequential consistency or SC}).
However, in both physically shared memory and PGAS, maintaining this sequentially consistent view of data among all concurrently accessing threads presents performance challenges.
The simplest way to maintain SC for a given data structure is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations. The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem magnifies.

Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
% One observation was that, when multiple threads of control concurrently access the same object, they can conduct themselves in one of two ways: contend or cooperate.
In the classic multithreading case, increasing concurrency increases contention and damages performance.
Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of potentially contended atomic operations, and retry if they fail. In both cases, as the number of concurrent accessors increases, the more they contend and the more failed synchronization operations there are.

At the core of a new synchronization technique called \emph{flat combining}~\cite{flatCombining} is the idea of having threads \emph{cooperate} rather than \emph{contend}.
Threads delegate their work to a single thread, which gives it the opportunity to combine multiple requests in data-structure specific ways and perform them free from contention.
Together, the reduced synchronization cost and combining allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.

>>>>

Globally shared data structures are one of the cornerstones of shared-memory and PGAS abstractions. 
In order for multiple concurrent threads to interact with one another correctly, they must observe shared data consistently. It is commonly accepted that the easiest consistency model to reason about enforces that all accesses appear to happen in some global serializable order (known as \emph{sequential consistency or SC}).
To preserve SC, operations on shared data structures should be \emph{linearizable}~\cite{herlihy1990linearizability}; that is, appear to happen atomically in some global total order.
However, in both physically shared memory and PGAS, maintaining linearizability presents performance challenges.
The simplest way is to have a single global lock that implements mutual exclusion and enforces a serializable order over read and write operations. The cost of literally serializing accesses in this way is typically considered prohibitive, even in physically shared memory.
With the massive amount of parallelism in a cluster of multiprocessors and with the increased cost of remote synchronization, the problem is magnified.

Much prior work has explored the scalability challenges of adding more concurrent accessors to shared data structures in shared memory systems.
% One observation was that, when multiple threads of control concurrently access the same object, they can conduct themselves in one of two ways: contend or cooperate.
In the classic multithreading case, increasing concurrency increases contention and damages performance.
Either all threads fight to obtain a single global lock, or in more sophisticated lock-free schemes, they attempt a series of contentious atomic operations, and retry if they fail. In both cases, as the number of concurrent accessors increases, the more they contend and the more synchronization operations fail.

At the core of a new synchronization technique called \emph{flat combining}~\cite{flatCombining} is the idea of having threads \emph{cooperate} rather than \emph{contend}.
Threads delegate their work to a single thread, giving it the opportunity to combine multiple requests in data-structure specific ways and performing them free from contention.
Together, the reduced synchronization cost and combining allow even a data structure with a single global lock to scale better than complicated concurrent data structures using fine-grained locking or lock-free mechanisms.
